---
layout: post
title: Getting Started With RGEE
date: 2024-01-15
tags: [ Google Earth Engine]
format:
  commonmark:
    variant: +yaml_metadata_block
    df-print: tibble
execute:
  echo: true
  message: false
  warning: false
  class-output: 'r'
---

```{r echo=FALSE}
options(
  max.print=100,
  vsc.use_httpgd=TRUE,
  device='quartz'
)
```
I need to get some data hosted through [Google Earth Engine](https://earthengine.google.com/). I've played around with the javacript and python libraries, but at heart I'm an R developer and would prefer to use earth engine in R if I can. 

This post is going to assume you have some python experience and can set up a python virtual environment of some sort. This is also done on a MacOS in zsh.  Getting a python environment setup is a little different on Windows. 

**Note:** I use VScode and call R using Radian a python package.  Because of this my default python environment was the environment that radian is in.  I basically could not change that environment no matter what I tried.  I would also note that it seems like many folks have trouble setting their environment within reticulate. 

## Setup a python environment and install rgee
To get rgee working we need to get a python environment set up.  To start from the terminal in bash or zsh (not an R terminal) make sure python is installed: 

```bash
❯ python --version
Python 3.12.1
```
I have python 3.12.1 installed, the latest versions as of the writing of this post. Then you can create a `venv` from whatever directory you wish (cd there first) and initialize it. 

```bash
python -m venv rgee_env
source rgee_env/bin/activate
```

You should have `(rgee_env)` in your terminal prompt if the environment was successfully created. Now we need to install the dependencies earthengine-api and Numpy. From the terminal again (not the R terminal): 

```bash
pip install earthengine-api
pip install numpy
```

You should have all the correct python libraries to run `rgee` from R.

## Setup Google Earth Engine Project

Create or use and existing account at the [Google Earth Engine](https://earthengine.google.com/) website and create a project.  Make note of the project ID. 

```{r}
library(rgee)
library(reticulate)
library(tidyverse)
library(terra)
```

## Login to Earth Engine
From here I am not going to evaluate any of these blocks because you cannot `ee_Initialize()` or get images that are stored together.  

Here we initialize my google cloud account and then make sure my credentials are good. 
```{r eval=FALSE}
ee_Initialize(user = 'mschmidty@gmail.com', drive = TRUE)
ee_check_credentials()
```

## Check Earth Engine and Python packages
Then we check that Earth Engine python packages are installed. 
```{r eval=FALSE}
ee_check()
ee_check_python_packages()
ee_check_packages()
```

## Make a Map
Next we tell Google Earth Engine what collection we would like to use, set a geometry and projection.  The geometry and projection are more so that I can download this image.
```{r eval=FALSE}
srtm <- ee$Image("USGS/SRTMGL1_003")
geometry <- ee$Geometry$Rectangle(
  coords = c(-108.274715, 37.076888, -107.213842, 37.965651),
  proj = "EPSG:4326",
  geodesic = FALSE
)
```

## Download the image as a `{terra}` SpatRast
This step downloads the above specified image to a folder on google drive. `ee_monitoring(task=as_rast)` tells us the status of the download while the next step `rast_col_rd<-columbine_rd|>ee_utils_future_value()` adds the image to memory. The easiest way to use the image from here is to save the raster to memory and then call it from there.  

```{r eval=FALSE}
columbine_rd <- ee_as_rast(
  image = srtm,
  via = "drive",
  region = geometry,
  lazy = TRUE
)
ee_monitoring(task=as_rast)
rast_col_rd<-columbine_rd|>ee_utils_future_value()
```
